---
title: "Three Tables of Genomics & Exploratory Data Analysis"
output: html_notebook
---
These notes come from Coursera's Statistics for Genomic Data Science course.
Three tables in genomics: Genomics Data, feature data and phenotype data.
Next, some exploratory data analysis will be conducted. 

#Three Tables in Genomics
```{r}
#Install Biobase package
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Biobase")
```
```{r}
library(Biobase)
```

Open the connection from bowtie experiment data (recount data) using a URL
```{r}
connection=url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=connection)
ls()
close(connection)
```

Change bodymap.eset to bm in order to shorten the variable name. Look at bodymap variable and we can see that it is made up of 19 samples and 5280 features. 
```{r}
bm=bodymap.eset
bm
```
```{r}
exp_data=exprs(bm) #Get expression data 
dim(exp_data) #Get dimensions of expression data
head(exp_data) #Print first couple of lines
```
Number of rows in pheno_data should be the same as the number of columns as expression data since each row in pheno data describes each of the columns in expression data. 
```{r}
pheno_data= pData(bm)#Wht are samples made up of, how are they collectd,etc
dim(pheno_data) 
```

```{r}
head(pheno_data)
```
Feature data describes genomic features of dataset
```{r}
feature_data=fData(bm)
dim(feature_data)
```
```{r}
head(feature_data)
```
###  Exploratory Data Analysis
Install and load necessary packages:
```{r}
install.packages('gplots')
```
```{r}
update.packages(repos='http://cran.rstudio.com/', ask=FALSE, checkBuilt=TRUE)
```

```{r}
install.packages("devtools")
```

```{r}
devtools::install_github('alyssafrazee/RSkittleBrewer')
```
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("org.Hs.eg.db")
```
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("AnnotationDbi")
```
```{r}
BiocManager::install("DBI")
```
```{r}
BiocManager::install("AnnotationDbi")
```
```{r}
BiocManager::install("org.Hs.eg.db")
```

```{r}
install.packages("org.Hs.eg.db")
```

```{r}
library("devtools")
```

```{r}
library("gplots")
```

```{r}
library("RSkittleBrewer")
```
```{r}
library("org.Hs.eg.db")
```
```{r}
library("AnnotationDbi")
```

Set the palette with defined colors.
```{r}
girly_colors=c('orchid3','magenta3','turquoise2','slateblue2','palegreen1')
palette(girly_colors)
```

Make sure circles are filled in when making plots.
```{r}
par(pch=19)
```

Check tables available
```{r}
ls()
```
```{r}
head(pheno_data)
```
### Explore the pheno_data table 
```{r}
table(pheno_data$gender)
```
```{r}
table(pheno_data$gender, pheno_data$race)
```

Based on the summary distribution is slightly skewed. 
```{r}
summary(exp_data)
```
```{r}
table(pheno_data$age)
```

To see NA values in the table:
```{r}
table(pheno_data$age, useNA='ifany')
```
Check if there are ay common missing value
```{r}
sum(pheno_data$age==" ", na.rm=TRUE)
```
Check to see if there are any missing values in the genomic data. Here wecheck row by row. 
```{r}
is.na(exp_data)[1,]
```
An alternativeis to sum the rows that have missing values and check for non-zero values:
```{r}
gene_na=rowSums(is.na(exp_data))
table(gene_na)
```
Check column by column alternatively:

```{r}
gene_na=colSums(is.na(exp_data))
table(gene_na)
```

### Plotting and Exploratory Data Analysis
```{r}
dim(exp_data)
```

```{r}
head(exp_data)
```

Here we see that most values for each of the samples are very close to zero. Next, we can log transform the data. 
```{r}
boxplot(exp_data[,1:19])
```
```{r}
boxplot(log2(exp_data+1), 
        col=1, 
        range=0,
        ylim=c(0,1)
        ,xlab="Sample",
        ylab="Expression",
        main="Boxplot of all samples expression data", font.main=1)
```
Make histogram. For this example, The histogram below represents the sample representedon coumn 1 of exp_data. 
```{r}
#par(mfrow=c(1,1))
hist(log2(exp_data[,1]+1),
     col=1,
     main="Histogram of sample ERS025098 expression data",
     xlab="Frequency",
     ylab="Counts", 
     font.main=1)

```
Density plots. We can ad plots on top of eachoter. Example with the first three samples:
```{r}
plot(density(log2(exp_data[,1]+1)),col=3, main="Density plots of first three samples")
lines(density(log2(exp_data[,2]+1)),col=1)
lines(density(log2(exp_data[,3]+1)),col=4)
```
qqplots. Each dot represents one quantile in each distribution. You can compare the percentiles sample by sample. Drawing 45 degree line helps to see the difference between the two samples. 
```{r}
qqplot(log2(exp_data[,1]+1),log2(exp_data[,2]+1), col=1, main="qqplot of samples 1 and 2 expression data",xlab="Sample 1",
     ylab="Sample 2")
abline(c(0,1))
```
Bland Altman Plot (MA Plot). This is done by taking the difference between the two samples as the y axis and the sum of the two samples as the x axis (both log transformed). In this plot we can see that as the counts increase the samples get closed together in similarity. 
```{r}
mm=log2(exp_data[,1]+1)-log2(exp_data[,2]+1)
aa=log2(exp_data[,1]+1)+log2(exp_data[,2]+1)
plot(aa, mm, col=4, main="Bland Altman Plot for samples 1 and 2")
```
For count based data,it is helpful to remove low-count features. Create a dataframe in order to be able to use the dplyr filtering commands to remove means less than one.
```{r}
install.packages('dplyr')
library('dplyr')
```

```{r}
exp_data=as.data.frame(exp_data)
filtered_edata=dplyr::filter(exp_data,rowMeans(exp_data)>1)
boxplot(as.matrix(log2(filtered_edata+1)),col=2, main="Filtered expression data for all samples")
```
### Compare dataset to annotation and metadat
Get the IDs for the features. 
```{r}
ls()
```
```{r}
head(pheno_data)
```
```{r}
IDS<-as.character(rownames(exp_data))
head(IDS)
```
Get the chromosome information for each of the ensemble IDs
```{r}
chr=AnnotationDbi::select(org.Hs.eg.db, keys=IDS,keytype="ENSEMBL",columns="CHR")
```
```{r}
head(chr)
```
Check that chr has the same dimensions as the expression dataset-it doesnt. Remove duplicated values from chromosome dataset. 
```{r}
dim(chr)
dim(exp_data)
```
```{r}
chr=chr[!duplicated(chr[,1]),]
dim(chr)
```
Check that all of the row names in the chromosome data match with the rownames in the expression data. 
```{r}
all(chr[,1]==rownames(exp_data))
```
Filter chromosome Y samples. Convert exp_data to dataframe to apply the filter command. 
```{r}

```

